Polynomial Chaos for the Estimation of Manning's Friction

### 2013-12-19 paper/intro.tex
work in progress

Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).  In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.  Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.  Recently, Sarri \emph{et al.} built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}. The implemented emulator was the Gaussian process
that requires using a combination of prior knowledge about the simulator and
appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
where they performed basic statistical and sensitivity analysis

>>>>

Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).  In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.  Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.  Recently, Sarri \emph{et al.} built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}. The implemented emulator was the Gaussian process
that requires using a combination of prior knowledge about the simulator and
appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
where they performed basic statistical and sensitivity analysis.
In another recent work, Mayo et el. reformulated a statistical data assimilation method generally used in the estimation of model states to estimate Manning’s n coefficients~\cite{Mayo2013}. 
They used a low-dimensional representations of Manning’s n coefficients and recovered it by assimilating
water elevation data.

### 2013-12-19 paper/tsunami.tex
work in progress

\alert{Need to describe how we handled the AMR in the UQ calculation.}

@@

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
\subsection{Parametric Uncertainty}
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $N$ by three different parameters
that are varied in three different regions but are assume
constant in each region. The three regions are as follows: 
on-shore who coefficnet is denoted by $N_1$, near-shore 
denoted by $N_2$, and deep-water denoted by $N_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.

>>>>

\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
at the gauges ready.... Not sure if this explanation is worth including here ///}

@@

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
\subsection{Parametric Uncertainty}
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $N$ by three different parameters
that are varied in three different regions but are assume
constant in each region. The three regions are as follows: 
on-shore who coefficnet is denoted by $N_1$, near-shore 
denoted by $N_2$, and deep-water denoted by $N_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $N$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $N \in [0.005-0.2]$.
